Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization

### 2024-06-27 manuscript.Rmd
remove the term "epistemic value"

This is the result of researchers applying one of two strategies to evade the self-imposed restrictions of preregistrations: writing a loose preregistration to begin with [@stefanBigLittleLies2022] or deviating from the preregistration afterward.
The latter is a frequent occurrence and, perhaps more worryingly, often remains undisclosed [@akkerEffectivenessPreregistrationPsychology2023; @claesenComparingDreamReality2021].
Both strategies may be used for sensible scientific reasons or with the self-serving intent of generating desirable results.
Thus, insisting on equating preregistration and confirmation has led to the criticism that, all things considered, preregistration is actually harmful and neither sufficient nor necessary for doing good science [@szollosiPreregistrationWorthwhile2020; @phamPreregistrationNeitherSufficient2021].

@@

# Epistemic value and the Bayesian rationale

Let us start by defining what we call expected epistemic value.
If researchers plan to conduct a study, they usually hope that it will change their assessment of some theory's verisimilitude [@niiniluotoVerisimilitudeThirdPeriod1998].
In other words, they hope to learn something from conducting the study.
The amount of knowledge researchers gain from a particular study concerning the verisimilitude of a specific theory is what we call epistemic value.
Researchers cannot know what exactly they will learn from a study before they run it.
However, they can develop an expectation that helps them decide about the specifics of a planned study.
This expectation is what we term expected epistemic value.
To make our three arguments, we must assume three things about what an ideal estimation process  entails and how it relates to what studies (preregistered vs not preregistered) to conduct.

1. Researchers judge the evidence for or against a hypothesis rationally.
2. They expect other researchers to apply a similar rational process.
2. Researchers try to maximize the expected epistemic value for other researchers.

The assumption of rationality can be connected to Bayesian reasoning and leads to our adoption of the framework.
Our rationale is as follows.
Researchers who decide to conduct a certain study are actually choosing a study to bet on.
They have to "place the bet" by conducting the study by investing resources and stand to gain epistemic value with some probability.
This conceptualization of choosing a study as a betting problem allows us to apply a "Dutch book" argument [@christensenCleverBookiesCoherent1991].
This argument states that any better must follow the axioms of probability to avoid being "irrational," i.e., accepting bets that lead to sure losses.
Fully developing a Dutch book argument for this problem requires careful consideration of what kind of studies to include as possible bets, defining a conversion rate from the stakes to the reward, and modeling what liberties researchers have in what studies to conduct.
Without deliberating these concepts further, we find it persuasive that researchers should not violate the axioms of probability if they have some expectation about what they stand to gain with some likelihood from conducting a study.
The axioms of probability are sufficient to derive the Bayes formula, on which we will heavily rely for our further arguments.
The argument is not sufficient, however, to warrant conceptualizing the kind of epistemic value we reason about in terms of posterior probability; that remains a leap of faith.
However, the argument applies to any reward function that satisfies the "statistical relevancy condition" [@wesleyc.salmonStatisticalExplanation1970; @fetzerStatisticalExplanations1974], that is, evidence only increases epistemic value for a theory if the evidence is more likely to be observed under the theory than under the alternative.
In particular, "diagnosticity" [@fiedlerWhatConstitutesStrong2017; @oberauerAddressingTheoryCrisis2019], a concept highlighted in recent psychological literature, seems to adhere to the statistical relevancy condition.

# Epistemic value and theoretical risk

Our first argument is that theoretical risk is crucial for judging evidential support for theories.
Put simply, risky predictions create persuasive evidence if they turn out to be correct.
This point is crucial because we attribute much of the appeal of a confirmatory research agenda to this notion.

@@

In short, we want to increase posterior probability $P(H|E)$.
Increases in posterior probability $P(H|E)$ are associated with increased epistemic value, of which we want to maximize the expectation.
So how can we increase posterior probability?
The Bayes formula yields three components that influence confirmation, namely $P(H)$, $P(E|H)$ and $P(E)$.
The first option leads us to the unsurprising conclusion that higher a priori probability $P(H)$ leads to higher posterior probability $P(H|E)$.
If a hypothesis is more probable to begin with, observing evidence in its favor will result in a hypothesis that is more strongly confirmed, all else being equal.
However, the prior probability of a hypothesis is nothing our study design can change.
The second option is equally reasonable; that is, an increase in $P(E|H)$ leads to a higher posterior probability $P(H|E)$.
$P(E|H)$ is the probability of obtaining evidence for a hypothesis when it holds.
We call this probability of detecting evidence, given that the hypothesis holds "detectability."
Consequently, researchers should ensure that their study design allows them to find evidence for their hypothesis, in case it is true.
When applied strictly within the bounds of null hypothesis testing, detectability is equivalent to power (or the complement of type-II error rate).
However, while detectability is of great importance for study design, it is not directly relevant to what a preregistration is comunicating to other researchers.
We later discuss how issues of detectability must be considered in a preregistration.
Thus, $P(E)$ remains to be considered.
Since $P(E)$ is the denominator, decreasing it can increase the posterior probability.
In other words, high risk, high reward.

@@

The last statement sounds like a truism but is directly related to Popper's seminal criterion of demarcation.
He stated that if it is impossible to prove that a hypothesis is false ($P(\neg{}E|\neg{}H) = 0$, theoretical risk is zero), it cannot be considered a scientific hypothesis [@popperLogicScientificDiscovery2002, p. 18].
We note these relations to underline that the Bayesian rationale we apply here is able to reconstruct many commonly held views on riskiness and epistemic value (but we of course differ from Popper on the central role of induction in science).

@@

Having discussed that increasing the theoretical risk will increase the epistemic value, it is intuitive to task preregistration with maximizing theoretical risk, i.e., a confirmatory research agenda.
Indeed, limiting the type-I error rate is commonly stated as  *the* central goal of preregistration [@nosekPreregistrationRevolution2018; @oberauerPreregistrationForkingPath2019; @rubinDoesPreregistrationImprove2020].
We argue that while such a conclusion is plausible, we must first consider at least two constraints that place an upper bound on the theoretical risk.

First, the theory itself limits theoretical risk:
Some theories simply do not make risky predictions, and preregistration will not change that.
Consider the case of a researcher contemplating the relation between two sets of variables.
Suppose each set is separately well studied, and strong theories tell the researcher how the variables within the set relate.
However, our imaginary researcher now considers the relation between these two sets.
For lack of a better theory, they assume that some relation between any variables of the two sets exists.
This is not a risky prediction to make in psychology [@orbenCrudReDefined2020].
However, we would consider it a success if the researcher would use the evidence from this rather exploratory study to develop a more precise (and therefore risky) theory, e.g., by using the results to specify which variables from one set relate to which variables from the other set, to what extent, in which direction, with which functional shape, etc., to be able to make riskier predictions in the future.
We will later show that preregistration increases the degree of belief in the further specified theory, though it remains low till being substantiated by testing the theory again.
This is because preregistration increases the expected epistemic value regardless of the theory being tested, as we will show.

Second, available resources limit theoretical risk.
Increasing theoretical risk $P(\neg{}E|\neg{}H)$ will usually decrease detectability $P(E|H)$ unless more resources are invested.
This is similar to the well known tradeoff between type-I error rate and statistical power.
Tasking preregistration with an increase in theoretical risk makes it difficult to balance this trade-off.
Mindlessly maximizing theoretical risk would either never produce evidence or require huge amounts of resources.
As noted before, we strive for high detectability and high theoretical risk in planning, conducting, and analyzing studies.
Maximizing one at the expense of the other is not necessarily beneficial for increasing epistemic value but depends on the specific function they apply to judge evidence and their specific location on the curve.
One advantage of our framework is that researchers can employ it to balance the trade-off more effectively assuming they are willing to make some simplifiying assumptions.

@@

Figure \ref{fig:posterior} shows exemplary the effect of theoretical risk (x-axis) on the posterior probability (y-axis) being certain (solid line) or uncertain (dashed line) about the theoretical risk of a study.
Our expectation of the gained epistemic value varies considerably depending on how uncertain we are about the theoretical risk a study took on.
Mathematically, uncertainty about theoretical risk is expressed through the variance (or rather entropy) of the distribution.
The increase in uncertainty (expressed as more entropic distributions) leads to a decreased expected epistemic value.

The argument for a confirmatory research agenda is that by increasing theoretical risk we increase expected epistemic value, i.e., moving to the right on the x-axis in Figure \ref{fig:posterior} increases posterior probability (on the y-axis).
However, if a hypothesis in a certain study has low theoretical risk, there is not much researchers can do about it.
However, studies do not only differ by how high the theoretical risk is but also by how certain the recipient is about the theoretical risk.
A study that has a very high theoretical risk (e.g., `r (1-confirmatory_tau)*100`% chance that if the hypothesis is wrong, evidence in its favor will be observed,) but has also maximum uncertainty will result in a posterior probability of `r round(confirmatory_uncertain*100)`%, while the same study with maximum certainty will result in `r round(confirmatory_certain*100)`% posterior probability.
The other factors (detectability, prior beliefs, measure of epistemic value) and, therefore, the extent of the benefit varies, of course, with the specifics of the study.
Crucially, even studies with some exploratory aspects benefit from preregistration, e.g., in this scenario with a $\tau$ = `r some_exploration_tau` (false positive rate of `r 1-some_exploration_tau`) moving from uncertain to certain increases the posterior from `r round(some_exploration_uncertain, 2)` to `r round(some_exploration_certain, 2)`.
We find it helpful to calculate an example because of the nonlinear nature of the evidence functions.

@@

1. Researchers judge the evidence for or against a hypothesis rationally.
2. They expect other researchers to apply a similar rational process.
2. Researchers try to maximize the expected epistemic value for other researchers.

@@

To summarize, we showed that both higher theoretical risk and lower uncertainty about theoretical risk lead to higher expected epistemic value across a variety of measures.
The former result that increasing theoretical risk leads to higher expected epistemic value reconstructs the appeal and central goal of preregistration of confirmatory research agendas.
However, theoretical risk is something researchers have only limited control over.
For example, theories are often vague and ill-defined, resources are limited, and increasing theoretical risk usually decreases detectability of a hypothesized effect (a special instance of this trade-off is the well-known tension between type-I error and statistical power).
While we believe that preregistration is always beneficial, it might be counterproductive to pursue high theoretical risk if the research context is inappropriate for strictly confirmatory research.
Specifically, appropriateness here entails the development of precise theories and the availability of necessary resources (often, large enough sample size, but also see @brandmaier2015lifespan) to adequately balance detectability against theoretical risk.

@@

Our latter result, on the importance of preregistration for minimizing uncertainty, has two important implications.
The first is, that even if all imaginable actions regarding promoting higher theoretical risk are taken, confirmatory research should be preregistered.
Otherwise, the uncertainty about the theoretical risk will diminish the advantage of confirmatory research.
Second, even under less-than-ideal circumstances for confirmatory research, preregistration is beneficial.
Preregistering exploratory studies increases the expected epistemic value by virtue of reducing uncertainty about theoretical risk.
Nevertheless, exploratory studies will have a lower expected epistemic value than a more confirmatory study if both are preregistered and have equal detectability.

@@

Second, researchers often have to deviate from the preregistration and make data-dependent decisions after the preregistration.
If the only goal of preregistration is to ensure confirmatory research, such changes are not justifiable.
However, under our rational, some changes may be justified.
Any change increases the uncertainty about the theoretical risk and may even decrease the theoretical risk.
The changes still may be worthwhile if the negative outcomes may be offset by an increase in detectability due to the change.
Consider a preregistration that failed to specify how to handle missing values, and researchers subsequently encountering missing values.
In such case, detectability becomes zero because the data cannot be analyzed without a post-hoc decision about how to handle the missing data.
Any such decision would constitute a deviation from the preregistration, which is possible under our proposed objective.
Note that a reader cannot rule out that the researchers leveraged the decision to decrease theoretical risk, i.e., picking among all options the one that delivers the most beneficial results for the theory (in the previous example, chosing between various options of handling missing values).
Whatever decision they make, increased uncertainty about the theoretical risk is inevitable and the expected epistemic value is decreased compared to a world where they anticipated the need to deal with missing data.
However, it is still justified to deviate.
After all they have not anticipated the case and are left with a detectablilty of zero.
Any decision will increase detectability to a non-zero value offsetting the increase in uncertainty.
The researchers also may do their best to argue that the deviation was not motivated by increasing theoretical risk, thereby, decreasing the uncertainty.
Ideally, there is a default decision that fits well with the theory or with the study design.
Or, if there is no obvious candidate, the researchers could conduct a multiverse analysis of the available options to deal with missings to show the influence of the decision [@steegenIncreasingTransparencyMultiverse2016].
In any case, deviations must be transparently reported and we applaud recent developments to standardize and normalize this process [@willrothBestLaidPlans2023].

@@

Contrary to what is widely assumed about preregistration, a preregistration is not necessarily a seal of confirmatory research.
Confirmatory research would almost always be less persuasive without preregistration, but in our view, preregistration primarily communicates the extent of confirmation, i.e., theoretical risk, of a study.
Clearly communicating theoretical risk is important because it reduces the uncertainty and hence increases expected epistemic value.

>>>>

This is the result of researchers applying one of two strategies to evade the self-imposed restrictions of preregistrations: writing a loose preregistration to begin with [@stefanBigLittleLies2022] or deviating from the preregistration afterward [@lakensWhenHowDeviate2024].
The latter is a frequent occurrence and, perhaps more worryingly, often remains undisclosed [@akkerEffectivenessPreregistrationPsychology2023; @claesenComparingDreamReality2021].
Both strategies may be used for sensible scientific reasons or with the self-serving intent of generating desirable results.
Thus, insisting on equating preregistration and confirmation has led to the criticism that, all things considered, preregistration is actually harmful and neither sufficient nor necessary for doing good science [@szollosiPreregistrationWorthwhile2020; @phamPreregistrationNeitherSufficient2021].

@@

# Persuation and the Bayesian rationale

If researchers plan to conduct a study, they usually hope that it will change their assessment of some theory's verisimilitude [@niiniluotoVerisimilitudeThirdPeriod1998].
Moreover, they hope to convince other researchers can be persuaded to change their believe in a theory as well.
Beforehand, researchers cannot know what evidence a study will provide but still must form an expectation in order to decide about the specifics of a planned study, including if they should preregister it.
If they can expect that preregistration helps them to persuade other researchers to change their believe, it is only rational to employ preregistration.
To make our three arguments, we must assume three things about what an ideal estimation process  entails and how it relates to what studies (preregistered vs not preregistered) to conduct.

1. Researchers judge the evidence for or against a hypothesis rationally.
2. They expect other researchers to apply a similar rational process.
2. Researchers try to maximize the expected persuasiveness for *other* researchers.

The assumption of rationality can be connected to Bayesian reasoning and leads to our adoption of the framework.
Our rationale is as follows.
Researchers who decide to conduct a certain study are actually choosing a study to bet on.
They have to "place the bet" by conducting the study by investing resources and stand to gain evidence for or against a theory with some probability.
This conceptualization of choosing a study as a betting problem allows us to apply a "Dutch book" argument [@christensenCleverBookiesCoherent1991].
This argument states that any better must follow the axioms of probability to avoid being "irrational," i.e., accepting bets that lead to sure losses.
Fully developing a Dutch book argument for this problem requires careful consideration of what kind of studies to include as possible bets, defining a conversion rate from the stakes to the reward, and modeling what liberties researchers have in what studies to conduct.
Without deliberating these concepts further, we find it reasonable that researchers should not violate the axioms of probability if they have some expectation about what they stand to gain with some likelihood from conducting a study.
The axioms of probability are sufficient to derive the Bayes formula, on which we will heavily rely for our further arguments.
The argument is not sufficient, however, to warrant conceptualizing persuasiveness in terms of posterior probability; that remains a leap of faith.
In fact, persuasiveness depends on how other researchers weigh evidence which differs between individuals.

However, the argument applies to any reward function that satisfies the "statistical relevancy condition" [@wesleyc.salmonStatisticalExplanation1970; @fetzerStatisticalExplanations1974], that is, evidence only increases believe for a theory if the evidence is more likely to be observed under the theory than under the alternative.
In particular, "diagnosticity" [@fiedlerWhatConstitutesStrong2017; @oberauerAddressingTheoryCrisis2019], a concept highlighted in recent psychological literature, seems to adhere to the statistical relevancy condition.

# Theoretical risk

Our first argument is that theoretical risk is crucial for judging the persuasiveness of evidence.
Put simply, risky predictions create persuasive evidence if they turn out to be correct.
This point is crucial because we attribute much of the appeal of a confirmatory research agenda to this notion.

@@

In short, we want to increase posterior probability $P(H|E)$.
Increases in posterior probability $P(H|E)$ are associated with increases in persuasiveness, of which we want to maximize the expectation.
So how can we increase posterior probability?
The Bayes formula yields three components that influence confirmation, namely $P(H)$, $P(E|H)$ and $P(E)$.
The first option leads us to the unsurprising conclusion that higher a priori probability $P(H)$ leads to higher posterior probability $P(H|E)$.
If a hypothesis is more probable to begin with, observing evidence in its favor will result in a hypothesis that is more strongly confirmed, all else being equal.
However, the prior probability of a hypothesis is nothing our study design can change.
The second option is equally reasonable; that is, an increase in $P(E|H)$ leads to a higher posterior probability $P(H|E)$.
$P(E|H)$ is the probability of obtaining evidence for a hypothesis when it holds.
We call this probability of detecting evidence, given that the hypothesis holds "detectability."
Consequently, researchers should ensure that their study design allows them to find evidence for their hypothesis, in case it is true.
When applied strictly within the bounds of null hypothesis testing, detectability is equivalent to power (or the complement of type-II error rate).
However, while detectability is of great importance for study design, it is not directly relevant to what a preregistration is communicating to other researchers.
We later discuss how issues of detectability must be considered in a preregistration.
Thus, $P(E)$ remains to be considered.
Since $P(E)$ is the denominator, decreasing it can increase the posterior probability.
In other words, high risk, high reward.

@@

The last statement sounds like a truism but is directly related to Popper's seminal criterion of demarcation.
He stated that if it is impossible to prove that a hypothesis is false ($P(\neg{}E|\neg{}H) = 0$, theoretical risk is zero), it cannot be considered a scientific hypothesis [@popperLogicScientificDiscovery2002, p. 18].
We note these relations to underline that the Bayesian rationale we apply here is able to reconstruct many commonly held views on how "risky" predictions are valued (but we of course differ from Popper on the central role of induction in science).

@@

Having discussed that increasing the theoretical risk will increase the persuasiveness, it is intuitive to task preregistration with maximizing theoretical risk, i.e., a confirmatory research agenda.
Indeed, limiting the type-I error rate is commonly stated as  *the* central goal of preregistration [@nosekPreregistrationRevolution2018; @oberauerPreregistrationForkingPath2019; @rubinDoesPreregistrationImprove2020].
We argue that while such a conclusion is plausible, we must first consider at least two constraints that place an upper bound on the theoretical risk.

First, the theory itself limits theoretical risk:
Some theories simply do not make risky predictions, and preregistration will not change that.
Consider the case of a researcher contemplating the relation between two sets of variables.
Suppose each set is separately well studied, and strong theories tell the researcher how the variables within the set relate.
However, our imaginary researcher now considers the relation between these two sets.
For lack of a better theory, they assume that some relation between any variables of the two sets exists.
This is not a risky prediction to make in psychology [@orbenCrudReDefined2020].
However, we would consider it a success if the researcher would use the evidence from this rather exploratory study to develop a more precise (and therefore risky) theory, e.g., by using the results to specify which variables from one set relate to which variables from the other set, to what extent, in which direction, with which functional shape, etc., to be able to make riskier predictions in the future.
We will later show that preregistration increases the degree of belief in the further specified theory, though it remains low till being substantiated by testing the theory again.
This is because preregistration increases the expected persuasiveness regardless of the theory being tested, as we will show.

Second, available resources limit theoretical risk.
Increasing theoretical risk $P(\neg{}E|\neg{}H)$ will usually decrease detectability $P(E|H)$ unless more resources are invested.
This is similar to the well known tradeoff between type-I error rate and statistical power.
Tasking preregistration with an increase in theoretical risk makes it difficult to balance this trade-off.
Mindlessly maximizing theoretical risk would either never produce evidence or require huge amounts of resources.
As noted before, we strive for high detectability and high theoretical risk in planning, conducting, and analyzing studies.
Maximizing one at the expense of the other is not necessarily beneficial for increasing persuasiveness but depends on the specific function they apply to judge evidence and their specific location on the curve.
One advantage of our framework is that researchers can employ it to balance the trade-off more effectively assuming they are willing to make some simplifiying assumptions.

@@

Figure \ref{fig:posterior} shows exemplary the effect of theoretical risk (x-axis) on the posterior probability (y-axis) being certain (solid line) or uncertain (dashed line) about the theoretical risk of a study.
Our expectation of the persuasiveness varies considerably depending on how uncertain we are about the theoretical risk a study took on.
Mathematically, uncertainty about theoretical risk is expressed through the variance (or rather entropy) of the distribution.
The increase in uncertainty (expressed as more entropic distributions) leads to a decreased expected persuasiveness.

The argument for a confirmatory research agenda is that by increasing theoretical risk we increase expected persuasiveness, i.e., moving to the right on the x-axis in Figure \ref{fig:posterior} increases posterior probability (on the y-axis).
However, if a hypothesis in a certain study has low theoretical risk, there is not much researchers can do about it.
However, studies do not only differ by how high the theoretical risk is but also by how certain the recipient is about the theoretical risk.
A study that has a very high theoretical risk (e.g., `r (1-confirmatory_tau)*100`% chance that if the hypothesis is wrong, evidence in its favor will be observed,) but has also maximum uncertainty will result in a posterior probability of `r round(confirmatory_uncertain*100)`%, while the same study with maximum certainty will result in `r round(confirmatory_certain*100)`% posterior probability.
The other factors (detectability, prior beliefs, measure of confirmation) and, therefore, the extent of the benefit varies, of course, with the specifics of the study.
Crucially, even studies with some exploratory aspects benefit from preregistration, e.g., in this scenario with a $\tau$ = `r some_exploration_tau` (false positive rate of `r 1-some_exploration_tau`) moving from uncertain to certain increases the posterior from `r round(some_exploration_uncertain, 2)` to `r round(some_exploration_certain, 2)`.
We find it helpful to calculate an example because of the nonlinear nature of the evidence functions.

@@

1. Researchers judge the evidence for or against a hypothesis rationally.
2. They expect other researchers to apply a similar rational process.
2. Researchers try to maximize the expected persuasiveness for other researchers.

@@

To summarize, we showed that both higher theoretical risk and lower uncertainty about theoretical risk lead to higher persuasiveness across a variety of measures.
The former result that increasing theoretical risk leads to higher expected persuasiveness reconstructs the appeal and central goal of preregistration of confirmatory research agendas.
However, theoretical risk is something researchers have only limited control over.
For example, theories are often vague and ill-defined, resources are limited, and increasing theoretical risk usually decreases detectability of a hypothesized effect (a special instance of this trade-off is the well-known tension between type-I error and statistical power).
While we believe that preregistration is always beneficial, it might be counterproductive to pursue high theoretical risk if the research context is inappropriate for strictly confirmatory research.
Specifically, appropriateness here entails the development of precise theories and the availability of necessary resources (often, large enough sample size, but also see @brandmaier2015lifespan) to adequately balance detectability against theoretical risk.

@@

Our latter result, on the importance of preregistration for minimizing uncertainty, has two important implications.
The first is, that even if all imaginable actions regarding promoting higher theoretical risk are taken, confirmatory research should be preregistered.
Otherwise, the uncertainty about the theoretical risk will diminish the advantage of confirmatory research.
Second, even under less-than-ideal circumstances for confirmatory research, preregistration is beneficial.
Preregistering exploratory studies increases the expected persuasiveness by virtue of reducing uncertainty about theoretical risk.
Nevertheless, exploratory studies will have a lower expected persuasiveness than a more confirmatory study if both are preregistered and have equal detectability.

@@

Second, researchers often have to deviate from the preregistration and make data-dependent decisions after the preregistration.
If the only goal of preregistration is to ensure confirmatory research, such changes are not justifiable.
However, under our rational, some changes may be justified.
Any change increases the uncertainty about the theoretical risk and may even decrease the theoretical risk.
The changes still may be worthwhile if the negative outcomes may be offset by an increase in detectability due to the change.
Consider a preregistration that failed to specify how to handle missing values, and researchers subsequently encountering missing values.
In such case, detectability becomes zero because the data cannot be analyzed without a post-hoc decision about how to handle the missing data.
Any such decision would constitute a deviation from the preregistration, which is possible under our proposed objective.
Note that a reader cannot rule out that the researchers leveraged the decision to decrease theoretical risk, i.e., picking among all options the one that delivers the most beneficial results for the theory (in the previous example, chosing between various options of handling missing values).
Whatever decision they make, increased uncertainty about the theoretical risk is inevitable and the expected persuasiveness is decreased compared to a world where they anticipated the need to deal with missing data.
However, it is still justified to deviate.
After all they have not anticipated the case and are left with a detectablilty of zero.
Any decision will increase detectability to a non-zero value offsetting the increase in uncertainty.
The researchers also may do their best to argue that the deviation was not motivated by increasing theoretical risk, thereby, decreasing the uncertainty.
Ideally, there is a default decision that fits well with the theory or with the study design.
Or, if there is no obvious candidate, the researchers could conduct a multiverse analysis of the available options to deal with missings to show the influence of the decision [@steegenIncreasingTransparencyMultiverse2016].
In any case, deviations must be transparently reported and we applaud recent developments to standardize and normalize this process [@willrothBestLaidPlans2023].

@@

Contrary to what is widely assumed about preregistration, a preregistration is not necessarily a seal of confirmatory research.
Confirmatory research would almost always be less persuasive without preregistration, but in our view, preregistration primarily communicates the extent of confirmation, i.e., theoretical risk, of a study.
Clearly communicating theoretical risk is important because it reduces the uncertainty and hence increases expected persuasiveness.
